---
title: "HELPmiss (Data Verbs) Activity"
author: "Mysha Nihal"
output: html_notebook
---

## Set Up:

```{r message=FALSE}
rm(list = ls())     # clean up your R environment

# load packages
library(tidyverse)  # includes lots of data verbs like `group_by()` and `summarise()`
library(mosaicData) # includes the `HELPmiss` data set 


# Load the `HELPmiss` data set into our RStudio environment
data("HELPmiss", package = "mosaicData")

```

## Task 1: Data Description

-   The HELP Study is a research project that examines the relationship between health behaviors, social support, and various aspects of life in a network of communities.

-   The HELPmiss dataset is derived from the HELP Study.

-   The dataset includes information on demographics, health behaviors, social support, and other relevant factors.

-   The specific individuals involved in the study can be found in the research paper or documentation.

-   The purpose of the study is to gain insights into how health behaviors and social support impact health outcomes and well-being.

-   Each case in the dataset represents an individual participant in the HELP Study.

## Task 2: Basic Summaries

1.  number of people (cases) in `HELPmiss` study

```{r}

number_of_people <- HELPmiss %>% 
  summarize(number_of_people = n())

```

2.  total number of times in the past 6 months entered a detox program (measured at baseline) for all the people in `HELPmiss`.

```{r}

total_detox_entries <- HELPmiss %>%
  summarize(total_detox_entries = sum(e2b, na.rm = TRUE))

```

3.  mean time (in days) to first use of any substance post-detox for all the people in `HELPmiss`

```{r}

mean_time_post_detox <- HELPmiss %>%
  summarize(mean_time_post_detox = mean(daysanysub, na.rm = TRUE))

```

## Task 3: Group Summaries

-   males versus females
-   homeless or not
-   substance
-   break down the homeless versus housed further, by sex
-   homeless versus housed broken down by substance

### males versus females

```{r}

people_gender <- HELPmiss %>%
group_by(sex) %>%
  summarize(people_gender = n())

```

```{r}

detox_entries_gender <- HELPmiss %>%
  group_by(sex) %>%
  summarize(detox_entries_gender = sum(e2b, na.rm = TRUE))

```

```{r}

mean_time_post_detox_gender <- HELPmiss %>%
  group_by(sex) %>%
  summarize(mean_time_post_detox_gender = mean(daysanysub, na.rm = TRUE))

```

### homeless or not

```{r}

people_gender_homeless <- HELPmiss %>%
group_by(homeless) %>%
  summarize(people_gender_homeless = n())

```

```{r}

detox_entries_homeless <- HELPmiss %>%
  group_by(homeless) %>%
  summarize(detox_entries_homeless = sum(e2b, na.rm = TRUE))

```

```{r}

mean_time_post_detox_homeless <- HELPmiss %>%
  group_by(homeless) %>%
  summarize(mean_time_post_detox_homeless = mean(daysanysub, na.rm = TRUE))

```

### substance

```{r}

people_gender_substance <- HELPmiss %>%
group_by(substance) %>%
  summarize(people_gender_substance = n())

```

```{r}

detox_entries_substance <- HELPmiss %>%
  group_by(substance) %>%
  summarize(detox_entries_substance = sum(e2b, na.rm = TRUE))

```

```{r}

mean_time_post_detox_substance <- HELPmiss %>%
  group_by(substance) %>%
  summarize(mean_time_post_detox_substance = mean(daysanysub, na.rm = TRUE))

```

### homeless versus housed broken down by sex

```{r}

people_gender_homeless_sex <- HELPmiss %>%
group_by(homeless, sex) %>%
  summarize(people_gender_homeless_sex = n())

```

```{r}

detox_entries_homeless_sex <- HELPmiss %>%
  group_by(homeless, sex) %>%
  summarize(detox_entries_homeless_sex = sum(e2b, na.rm = TRUE))

```

```{r}

mean_time_post_detox_homeless_sex <- HELPmiss %>%
  group_by(homeless, sex) %>%
  summarize(mean_time_post_detox_homeless_sex = mean(daysanysub, na.rm = TRUE))

```

### homeless versus housed broken down by substance

```{r}

people_gender_homeless_substance <- HELPmiss %>%
group_by(homeless, substance) %>%
  summarize(people_gender_homeless_substance = n())

```

```{r}

detox_entries_homeless_substance <- HELPmiss %>%
  group_by(homeless, substance) %>%
  summarize(detox_entries_homeless_substance = sum(e2b, na.rm = TRUE))

```

```{r}

mean_time_post_detox_homeless_substance <- HELPmiss %>%
  group_by(homeless, substance) %>%
  summarize(mean_time_post_detox_homeless_substance = mean(daysanysub, na.rm = TRUE))

```

## Task 4: Data Visualization & Observations

```{r}

ggplot(HELPmiss, aes(x = age, y = daysanysub, color = substance)) +
  geom_point() +
  labs(x = "Age (years)", y = "Days to 1st post-detox substance use", color = "Substance") +
  theme_minimal()

```
